\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-02-16 jasa/paper-jasa.tex

In light of this result, the likelihood ratio estimation problem can now
be recast as a (probabilistic) classification problem, by noticing that the decision
function
\begin{equation}\label{eqn:best-s-clf}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
\end{equation}
modeled by a classifier trained to distinguish samples $\mathbf{x} \sim p_{\theta_0}$
from samples $\mathbf{x} \sim p_{\theta_1}$ satisfies conditions of
Theorem~\ref{thm:ratio-equivalence} (see Appendix~\ref{app:clf-for-s} for further details).
In other words, supervised learning yields a sufficient
procedure for Theorem~\ref{thm:ratio-equivalence} to hold, guaranteeing that any
{\it universally strongly consistent} algorithm can be used for learning $s^*$.
Note however, that it is not a necessary procedure since
Theorem~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
density ratio, i.e., not for $m(r(\mathbf{x})) = \frac{1}{1 +
r(\mathbf{x})}$ only.
Equivalently,
Theorem~\ref{thm:ratio-equivalence} shows that in case we learn a probabilistic
classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modeling
$p_{\mathbf{U}}(u=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.

@@

As outlined above, any consistent probabilistic classification algorithm can be
used for learning an approximate  map $\hat{s}(\mathbf{x})$ of
Eqn.~\ref{eqn:best-s-clf}. In the common case where the density ratio is
expected to smoothly vary around $\mathbf{x}$, we would however recommend
learning models whose output value $\hat{s}(\mathbf{x})$ also smoothly varies
around $\mathbf{x}$, such as neural networks. For small training sets, tree-based
methods are not expected to work so well for this use case, since they usually
model $\hat{s}(\mathbf{x})$ as a non-strictly monotonic composition of step functions.
In particular, in such cases where $s(\mathbf{x})$ is not monotonic with $r(\mathbf{x})$,
the induced probability does not factorize as in Eqn.~\ref{eqn:factorization}, resulting in a biased
approximation of the density ratio.

@@

Using the method outlined in Section~\ref{sec:approx},
figures~\ref{fig:2a}, \ref{fig:2b} and \ref{fig:2c} show the approximated
ratio when using respectively a linear model, a 2-layer neural network or a random
forest. The blue curve corresponds to the true target log-ratio, the green curve is
the direct approximation $\log \left( \frac{\hat s(\mathbf{x})}{1 - \hat s(\mathbf{x})} \right)$
without calibration, while the red curve is the approximation $\log \frac{\hat p(\hat
s(\mathbf{x}) | \theta_0)}{\hat p(\hat s(\mathbf{x}) | \theta_1)}$ as calibrated
using histograms. Finally, the
cyan curve shows the approximated ratio when decomposing the mixtures, as
described in Section~\ref{sec:mixture}.
Hyper-parameters of the classifiers are all tuned on a finite grid of values
in order to minimize the mean squared error, as estimated by 3-fold cross-validation.

@@

The first fact to highlight is that, in the finite sample setting, uncalibrated
direct approximations $\log \left( \frac{\hat s(\mathbf{x})}{1 - \hat
s(\mathbf{x})} \right)$, as shown in green, do not yield very accurate estimates
of the ratio, even when using universal approximators such as a neural network
or a random forest. This observation is at the origin of this work, and led us
to show that the ratio $r(\mathbf{x})$ can still be computed exactly by forming
the equivalent statistic
$\frac{p(s(\mathbf{x})|\theta_0)}{p(s(\mathbf{x})|\theta_1)}$, provided the
transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
constructing $\frac{\hat p(\hat s(\mathbf{x})|\theta_0)}{\hat p(\hat
s(\mathbf{x})|\theta_1)}$ as we propose, estimates appear much closer to the
true ratio than their direct counterparts, as illustrated in red. The
approximated ratios show to be accurate for all three classifiers, in particular in regions with enough density
(i.e., from $\mathbf{x}=-3$ to $\mathbf{x}=2$) and where the capacity of the
classifier is focused. By contrast, in regions where the densities are low
(i.e., at the boundaries) the approximated ratio show to have high variance
since the underlying classifier cannot make very accurate predictions, because
of the few training samples it is trained on in these parts of the input space.
Accordingly, when leveraging the fact that densities are mixtures and decomposing their
ratio, the capacity of the underlying classifiers can be focused on easier
supervised learning tasks, resulting as expected in even more accurate approximations, as
shown in cyan for all three classifiers.

As shown in the figures, qualitatively different results are obtained depending
the classification algorithm embedded within the approximation of the ratio. The
neural network-based approximation appears to work well, due to the capacity of
the classifier to model a decision function $s(\mathbf{x})$ continuous and
almost monotonic with the ratio. The random forest yields a decision function
which is piece-wise constant, therefore non-continuous and not strictly
monotonic with the ratio in the finite sample regime. On this problem, the
random forest still approximates the ratio correctly, but more variance can be
observed. Finally, the linear model produces a decision function that is
continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
the ratio around $\mathbf{x}=1$, this still results in an accurate
approximation, but poorer results should be expected on more complex problems.
Similarly, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is also key to obtain accurate results. While standard
histograms with fixed bins have been used here for illustrative purposes, we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis.

@@

\glnote{Discuss diagnostics?}
\glnote{Better highlight advantages of the method?}
\glnote{Conduct more thorough and convincing experiments?}

>>>>

In light of this result, the likelihood ratio estimation problem can now
be recast as a (probabilistic) classification problem, by noticing that the decision
function
\begin{equation}\label{eqn:best-s-clf}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
\end{equation}
modeled by a classifier trained to distinguish samples $\mathbf{x} \sim p_{\theta_0}$
from samples $\mathbf{x} \sim p_{\theta_1}$ satisfies conditions of
Theorem~\ref{thm:ratio-equivalence} (see Appendix~\ref{app:clf-for-s} for further details).
In other words, supervised learning yields a sufficient
procedure for Theorem~\ref{thm:ratio-equivalence} to hold, guaranteeing that any
{\it universally strongly consistent} algorithm can be used for learning $s^*$.
Note however, that it is not a necessary procedure since
Theorem~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
density ratio, i.e., not for $m(r(\mathbf{x})) = \frac{1}{1 +
r(\mathbf{x})}$ only.
Equivalently,
Theorem~\ref{thm:ratio-equivalence} shows that in case we learn a probabilistic
classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modeling
$p_{\mathbf{U}}(u=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.
For this reason, the proposed method is expected to be more robust than
directly using $\frac{1 - s(\mathbf{x})}{s(\mathbf{x})}$ as an approximate of
$r(\mathbf{x})$.

@@

As outlined above, any consistent probabilistic classification algorithm can be
used for learning an approximate  map $\hat{s}(\mathbf{x})$ of
Eqn.~\ref{eqn:best-s-clf}. In the common case where the density ratio is
expected to smoothly vary around $\mathbf{x}$, we would however recommend
learning models whose output value $\hat{s}(\mathbf{x})$ also smoothly varies
around $\mathbf{x}$, such as neural networks. For small training sets, tree-based
methods are not expected to work so well for this use case, since they usually
model $\hat{s}(\mathbf{x})$ as a non-strictly monotonic composition of step functions.
In particular, in such cases where $\hat s(\mathbf{x})$ is not monotonic with $r(\mathbf{x})$,
the induced probability does not factorize as in Eqn.~\ref{eqn:factorization}, resulting in a biased
approximation of the density ratio.

@@

Using the method outlined in Section~\ref{sec:approx},
figures~\ref{fig:2a}, \ref{fig:2b} and \ref{fig:2c} show the approximated
ratio when using respectively a linear model, a 2-layer neural network and a random
forest. The blue curve corresponds to the true target log-ratio, the green curve is
the direct approximation $\log \left( \frac{1 - \hat s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$
without calibration, while the red curve is the approximation $\log \frac{\hat p(\hat
s(\mathbf{x}) | \theta_0)}{\hat p(\hat s(\mathbf{x}) | \theta_1)}$ as calibrated
using histograms. Finally, the
cyan curve shows the approximated ratio when decomposing the mixtures, as
described in Section~\ref{sec:mixture}.
Hyper-parameters of the classifiers are all tuned on a finite grid of values
in order to minimize the mean squared error, as estimated by 3-fold cross-validation.

@@

The first fact to highlight is that, in the finite sample setting, uncalibrated
direct approximations $\log \left( \frac{1 - \hat
s(\mathbf{x})}{\hat s(\mathbf{x})} \right)$, as shown in green, do not yield very accurate estimates
of the ratio, even when using universal approximators such as a neural network
or a random forest. This observation is at the origin of this work, and led us
to show that the ratio $r(\mathbf{x})$ can still be computed exactly by forming
the equivalent statistic
$\frac{p(s(\mathbf{x})|\theta_0)}{p(s(\mathbf{x})|\theta_1)}$, provided the
transformation $s(\mathbf{x})$ is monotonic with $r(\mathbf{x})$. Indeed, when
constructing $\frac{\hat p(\hat s(\mathbf{x})|\theta_0)}{\hat p(\hat
s(\mathbf{x})|\theta_1)}$ as we propose, estimates appear much closer to the
true ratio than their direct counterparts, as illustrated in red. The
approximated ratios show to be accurate for all three classifiers, in particular in regions with enough density
(i.e., from $\mathbf{x}=-3$ to $\mathbf{x}=2$) and where the capacity of the
classifier is focused. By contrast, in regions where the densities are low
(i.e., at the boundaries) the approximated ratio show to have high variance
since the underlying classifier cannot make very accurate predictions, because
of the few training samples it is trained on in these parts of the input space.
Accordingly, when leveraging the fact that densities are mixtures and decomposing their
ratio, the capacity of the underlying classifiers can be focused on easier
supervised learning tasks, resulting as expected in even more accurate approximations, as
shown in cyan for all three classifiers.

As shown in the figures, qualitatively different results are obtained depending
the classification algorithm embedded within the approximation of the ratio. The
neural network-based approximation appears to work well, due to the capacity of
the classifier to model a smooth decision function $s(\mathbf{x})$
almost monotonic with the ratio. The random forest yields a decision function
which is piece-wise constant, therefore non-continuous and not strictly
monotonic with the ratio in the finite sample regime. On this problem, the
random forest still approximates the ratio correctly, but more variance can be
observed. Finally, the linear model produces a decision function that is
smooth and continuous but only piece-wise monotonic with the ratio. Due to the symmetry of
the ratio around $\mathbf{x}=1$, this still results in an accurate
approximation, but poorer results should be expected on more complex problems.
Similarly, calibrating $\hat s(\mathbf{x})$ to construct $\hat p(\hat
s(\mathbf{x}))$ is also key to obtain accurate results. While standard
histograms with fixed bins have been used here for illustrative purposes, we
anticipate that the embedded density estimation algorithm should be carefully
chosen and tuned, on a case by case basis.

@@

\subsection{Advanced example}

\glnote{todo}
